---
title: "Reproducible Research Course Project 1"
author: "Joshua Stein"
date: "2024-03-27"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction

In this document, I will download a dataset provided by Professor Roger Peng as part of the Reproducible Research Class on Coursera. The dataset describes activity monitoring data from personal fitness monitoring devices. In this document, the datset will be loaded and processed in RStudio, and several questions regarding the dataset will be investigated. 

# Loading and Processing the Data

The data was first loaded into RStudio and processed for proper analysis. 

```{r input}
url <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip"
download.file(url, destfile = "/Users/joshuastein/Desktop/hello-world/ReproducibleResearch/CourseProject1/data.zip", mode = "wb")

unzip("/Users/joshuastein/Desktop/hello-world/ReproducibleResearch/CourseProject1/data.zip")

data <- read.csv("activity.csv", na.strings = "NA")
data$date <- as.Date(data$date)
```

# How many steps were taken each day?

The cut and aggregate functions were used to determine the number of steps taken per day. 

```{r steps}
dateintervals <- cut(data$date, breaks = "day")
stepsbydate <- aggregate(steps ~ dateintervals, data = data, FUN = sum, na.action = na.pass)
```

The number of steps per day seems to widely variable from below 50 to over 20,000 steps per day.

```{r table, results='asis'}
library(knitr)
kable(stepsbydate)
```

The mean and median were also calculated: 

```{r summary}
mean(stepsbydate$step, na.rm = TRUE)
median(stepsbydate$steps, na.rm = TRUE)
```

The data was also graphed on a histogram for visualization. The majority of total steps seem to fluctuate between 10,000 and 15,000 steps per day: 
```{r histogram}
library(ggplot2)
g <- ggplot(data = stepsbydate, aes(x = steps)) + geom_histogram()
print(g)
```

# What is the average daily activity pattern?

It's apparent that the daily number of steps widely varies. In this section, we investigated potential patterns to daily activity by creating a Time Series Plot and looking at 5-minute interval:

```{r avgpattern}
patternandsteps <- cut(data$date, breaks = "day")
stepsbyinterval <- aggregate(steps ~ interval, data = data, FUN = mean)
plot(stepsbyinterval$interval, stepsbyinterval$steps, type = "l", xlab = "5 minute interval", ylab = "Average Number of Steps", main = "Time Series Plot")
```

The 5-minute interval with the highest average number of steps is 835: 

```{r maxsteps}
indexnum <- which.max(stepsbyinterval$steps)
print(stepsbyinterval$interval[indexnum])
```

# Input missing values

In this section, we wanted to investigate whether the NAs in the current dataset influences the results. First, let's determine the number of NAs:

```{r NAnum}
sum(is.na(data$steps))
sum(is.na(data$interval))
```

Now, let's replace the NA value with mean of steps from that day using a while loop. The data with no NAs is then written to a csv file. 

```{r NAreplace}
stepsbydate$dateintervals <- as.Date(stepsbydate$dateintervals)
i <- 1
while (i <= length(data$steps)){
  if(is.na(data$steps[i])){
    initdate <- as.Date("2012-10-01")
    whatisthedate <- data$date[i]
    indexnumdays <- difftime(whatisthedate, initdate, units = "days")
    
    if(indexnumdays < 1 || indexnumdays > length(stepsbydate$steps) || is.na(stepsbydate$steps[i])){
      data$steps[i] <- 0
    } 
    else {
      data$steps[i] <- stepsbydate$steps[indexnumdays]    
    }
  }
  i <- i + 1
}
write.csv(data, file = "activitynoNA.csv", row.names = TRUE)

sum(is.na(data$steps))
sum(is.na(data$interval))
```

There is no more NAs in the dataset. Now, let's reanalyze the dataset. First, let's process the data again.

```{r repeat}
dateintervalsnoNA <- cut(data$date, breaks = "day")
stepsbydatenoNA <- aggregate(steps ~ dateintervals, data = data, FUN = sum, na.action = na.pass)
kable(stepsbydatenoNA)
```

The NA values, observed on 10/1 and 10/8, have been replaced with 0 indicating the code worked as expected. To confirm, let's determine the mean and median:

```{r summaryrepeat}
mean(stepsbydatenoNA$step, na.rm = TRUE)
median(stepsbydatenoNA$steps, na.rm = TRUE)
```

There appears to be a significant change in the summary statistics. The mean of our dataset with no NAs is 9354.23 (compared to 10766.19 in the dataset with NAs). The median for the NA-free dataset is 10395 (compared to 10765 in the dataset with NAs). We expect the mean to change signficantly since we have changed many data values from NA to 0. The additon of the 0s have skewed the mean to about ~1500 units below that of the dataset with the NAs. Since we didn't add any new values, we don't see any change in the median. Let's now visualize this using a histogram:

```{r histogramrepeat}
library(ggplot2)
g <- ggplot(data = stepsbydatenoNA, aes(x = steps)) + geom_histogram()
print(g)
```

We see a huge spike in the number of data values at 0. The rest of the analysis seems to be identical. From this dataset, we can confidently conclude that all NAs have been replaced with the mean number of steps for that day or replaced with a 0 for those days (i.e. 10/1 and 10/8) where there is no available mean.

# Are there differences in activity patterns between weekdays and weekends?

First, let's add a new variable describing the type of day (weekend vs weekday).

```{r dayorend}
library(dplyr)

datawithNAwithday <- mutate(data, dayname = weekdays(as.Date(dateintervals)))
datawithNAwithdaytype <- mutate(datawithNAwithday, daytype = ifelse(dayname == "Saturday" | dayname == "Sunday", "weekend", "weekday"))
datafinal <- aggregate(steps ~ interval + daytype, datawithNAwithdaytype, mean)
```

And we can graph using ggplot.

```{r graphagain}

library(ggplot2)
g <- ggplot(datafinal, aes(x = interval, y = steps, color = daytype)) +
  geom_line() +
  facet_wrap(~daytype) + 
  labs(title = "5-minute Interval with Total Number of Steps Based on Type of Day",
       x = "5-minute Interval",
       y = "Number of Steps") 
print(g)
```